---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
unzip("activity.zip")
data <- read.csv("activity.csv", colClasses = c("numeric", "character", "numeric"))
```

## What is mean total number of steps taken per day?

```{r}
library(dplyr)
##groups the data by date
data2 <- group_by(data, date)

##calculate total number of steps per day
stepsperday <- summarize(data2, sum(steps, na.rm=TRUE))
stepsperday

##assigns a convenient column name to total number of steps 
colnames(stepsperday)[2] <- "steps"

## creates an hitstogram of total number of steps per day
hist(stepsperday$steps)

##mean total number of steps taken per day
x <- mean(stepsperday$steps)
x
## median of the total number of steps taken per day
y <- median(stepsperday$steps)
y
```
The mean number of steps taken per day is `r x`. The median number of steps taken per day is `r y`

## What is the average daily activity pattern?

```{r}
##calculate mean number of steps per 5-min interval averaged across all days
data3 <- group_by(data, interval)
stepsperinterval <- summarize (data3, mean(steps, na.rm = TRUE))

## plots the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
plot(stepsperinterval, type="l")
```

```{r}
## Finds which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps
colnames(stepsperinterval)[2] <- "stepsavg"
data4 <- filter(stepsperinterval, stepsavg == max(stepsperinterval$stepsavg))
data4
z <- data4[1,1]
```
The interval `r z` contains the maximum number of steps. 

## Imputing missing values

```{r}
## number of NAs for steps variable
n1 <- length(which(is.na(data$steps)))
n1
```
There are `r n1` missing values in the dataset.

We will replace missing values by the mean number of steps across days for the corresponding interval.

The steps are as follow:

1) Merge the original dataset with the summary of mean number of steps across days for each interval.This way, each row will include the value observed on that particular interval for a particular day in the steps variable, as well as the mean number of steps across days for that particular interval in the stepsavg variable

2) Whevener a row contains NA in the steps column, it will be replaced by the value from the stepsavg column

3) The stepsavg column is removed, and columns and observations are re-ordered to match the original dataset.

```{r}
##adds a column containing the mean number of steps across days for corresponding interval to each row
data5 <- merge(data, stepsperinterval, by = "interval")
##NAs for the steps variable are replaced by mean value across day for corresponding interval
data5$steps[is.na(data5$steps)] <- data5$stepsavg[is.na(data5$steps)]
##creates a dataset equal to the original one, with missing data filled in.
data5 <- select(data5, -stepsavg)
data5 <- arrange(data5, date, interval)
data5 <- select(data5, steps, date, interval)
head(data5,20)
##checks that the dataset contains no NA
length(which(is.na(data5$steps)))
```

We want to check the effect of these new values.
```{r}
##groups the data by date
data6 <- group_by(data5, date)

##calculate total number of steps per day
stepsperday2 <- summarize(data6, sum(steps, na.rm=TRUE))
stepsperday2

##assigns a convenient column name to total number of steps 
colnames(stepsperday2)[2] <- "steps"

## creates an hitstogram of total number of steps per day
hist(stepsperday2$steps)

##mean total number of steps taken per day
x2 <- mean(stepsperday2$steps)
x2
## median of the total number of steps taken per day
y2 <- median(stepsperday2$steps)
y2
```
The mean number of steps taken per day with the new data is `r x2`. The median number of steps taken per day with the new data is `r y2`. This increases the mean of total number of steps taken per day, as well as the median (to a lesser extent). Interestingly, the mean and the median are now equal. If you are good at stats, maybe you can explain me that in the comments/feedback section?(Thanks!)

## Are there differences in activity patterns between weekdays and weekends?

```{r}
##takes the dataset with added data instead of NAs, and converts data into POSIXlt format
data5$date <- strptime(data5$date, format="%Y-%m-%d")

##creates the weekday column, containing a numeric value from 1(Monday) to 7(Sunday)
data5$weekday <- format(data5$date, format = "%u")
data5$weekday <- as.numeric(data5$weekday)

##Weekday values lesser than or equal to five are renamed "weekday"
data5$weekday[data5$weekday <= 5] <- "weekday"
##Remaining values ("6" and "7" are renamed "weekend")
data5$weekday[data5$weekday == "6"|data5$weekday == "7"] <- "weekend"
head(data5, 20)
## values from column weekday are coerced into factors
data5$weekday <- as.factor(data5$weekday)
class(data5$weekday)
```

In the next chunk of code, we want to mkake a panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
data8 <- group_by(data8, weekday, interval)
data9 <- summarize(data8, mean(steps))
colnames(data9)[3] <- "steps"

## needs a tad more work to make it look good
qplot(x= interval, y= steps, data=data9,geom=c("line"),facets=.~weekday)        
```

